\documentclass[11pt,a4paper]{article}

% ── Packages ──────────────────────────────────────────────────────────
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{xcolor}
\usepackage{enumitem}
\usepackage{subcaption}
\usepackage{bm}

% ── Macros ────────────────────────────────────────────────────────────
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\argmin}{\operatorname*{arg\,min}}
\newcommand{\argmax}{\operatorname*{arg\,max}}
\DeclareMathOperator*{\diag}{diag}
\newcommand{\bestk}{\textrm{best}@k}

% ── Title ─────────────────────────────────────────────────────────────
\title{Matryoshka Reward Policy Optimization:\\
Structuring Reward So Diversity Survives for Test-Time Search\\[6pt]
{\large Draft --- targeting an oral submission}}
\author{Terminal Contrastive RL Project}
\date{June 2026}

\begin{document}
\maketitle

\begin{abstract}
LLM post-training increasingly feeds an inference-time search that draws many
candidates and keeps the best.  In this regime the search does the exploiting,
so training should preserve the diversity that search consumes---yet scalar
reinforcement learning sharpens the policy onto a single reward mode and
destroys it.  We argue that reward post-training is better understood as
\emph{shaping} a policy's diversity for downstream search than as maximising a
scalar, and we study how the \emph{structure} of the reward signal governs that
diversity.  We show that under outcome-level policy gradients a vector-valued
reward can be preserved at the gradient level essentially for free: every
per-axis gradient lies in the low-dimensional span of a group's per-trajectory
gradients, so combining axes---including by min-norm (MGDA) methods---is a cheap
reweighting rather than extra backward passes.  Building on this we propose
\textbf{Matryoshka Reward Decomposition}, which learns a reward code with a few
interpretable, quality-aligned dimensions and a deliberately entangled
remainder, supplying downstream search with controllable rather than undirected
diversity.

Empirically, across an agentic tool-use task and standard benchmarks (GSM8K,
MBPP code generation, Wordle), optimising for inter-sample agreement collapses
tool use and search coverage, and we identify the governing condition: reward
axes must be both non-collinear \emph{and} aligned with task
quality---non-collinearity alone is not enough.  We further find that min-norm
(MGDA) gradient combination is substantially more \emph{reliable} across random
seeds than fixed or random scalarisation while matching its mean, and that it
avoids the consensus collapse the other combiners suffer---except in a
degenerate floor regime (Wordle), which sharpens rather than contradicts the
condition.  The takeaway: reliability, not mean reward, is the axis on which
gradient-level vector methods earn their keep, and the benefit tracks reward-axis
structure as predicted.  The reliability advantage holds from 1.5B to 7B---the
verifier-free baselines collapse harder at scale while MGDA preserves base
performance---so it is not a small-model artifact.
\end{abstract}

\tableofcontents
\newpage

% ======================================================================
\section{Motivation: Search Wants Diversity; Scalar Rewards Destroy It}
\label{sec:motivation}

\subsection{The exploration/exploitation split has moved}

LLMs are increasingly deployed \emph{inside} an inference-time search loop:
sample $k$ candidates and keep the best under some selector (a verifier, a
reward model, a unit test, or an evolutionary outer loop).  In this regime
the search does the exploiting.  What training must supply is
\emph{coverage}: a policy whose samples spread across enough distinct,
high-quality strategies that the selector has something to select from.

This regime is not hypothetical.  Code-engineering agents such as AIDE
(Jiang et al.\ 2025) hill-climb a metric by tree-searching the space of
candidate solutions---draft, run, read the feedback, propose the next
edit---and beat the best \emph{linear} (single-pass) agent by roughly
$4\times$ on MLE-bench.  The lesson transfers directly: the search is the
exploiter, and its bottleneck is the proposer's ability to keep generating
\emph{distinct, useful} candidates as the budget is spent.  A policy that has
been sharpened onto one mode starves that search.  Our experiments
(Section~\ref{sec:exp}) measure this bottleneck directly.

Standard policy-gradient post-training optimises a fixed scalar reward:
\begin{equation}
  \nabla J(\theta) \;=\; \E_{\tau \sim \pi_\theta}\!\Big[
    \nabla \log \pi_\theta(\tau) \;\cdot\; A(\tau)
  \Big],
  \qquad A(\tau) \in \R.
  \label{eq:pg-scalar}
\end{equation}
Maximising a scalar sharpens the distribution onto whichever single
strategy tops that scalar.  The entropy drops, every extra sample becomes a
near-duplicate, and the diversity that downstream search runs on collapses
precisely when it is most valuable.  This is the failure mode our
experiments and others' (Section~\ref{sec:related}) repeatedly hit: $\bestk$
stops improving with $k$ because the $k$ draws are no longer distinct.

\subsection{Scalarising a privileged teacher throws away structure}

The same scalar bottleneck has a second cost.  When a \emph{privileged
teacher} returns a vector of feedback signals $\bm{R}(\tau) \in \R^K$, the
conventional pipeline reduces it to a scalar via a fixed or random weighted
sum \emph{before} computing the gradient.  This is informationally lossy: a
trajectory that is right on $K{-}1$ axes but wrong on one gets the same
scalar as a trajectory that is mediocre on all axes.

\paragraph{Example.}
Consider a bash-agent rollout that produces the correct answer but uses an
invalid tool-call format.  A 5-axis teacher returns
\[
  \bm{A}(\tau) = (\underbrace{+1}_{\text{correct}},\;
                   \underbrace{+1}_{\text{reasoning}},\;
                   \underbrace{-1}_{\text{tool format}},\;
                   \underbrace{+1}_{\text{state match}},\;
                   \underbrace{-0.5}_{\text{verbose}}).
\]
Scalarising with equal weights gives $A = +0.3$---``this trajectory was
$0.3$ good.''  The gradient says ``do more of this, a bit,'' and the model
drifts.  A vector-aware update can instead say ``fix the tool call; the rest
is fine,'' \emph{and} can route the correct-but-verbose and
correct-but-different-state variants into distinct regions of policy space
so search later sees both.  That targeted, diversity-preserving signal is
what pedagogical RL is for, and scalarisation destroys it.

\subsection{One thesis}

These are the same problem viewed from two sides.  Scalarisation collapses
the policy (bad for search) \emph{and} collapses the teacher's structure
(bad for credit assignment).  Our claim is that the right object to
optimise is not a scalar reward but a \emph{structured reward code}, and
that the value of doing so is governed by a single, measurable quantity:
how much non-collinear information that code actually carries.  The rest of
the paper makes ``structured reward code'' precise (Section~\ref{sec:mrd}),
shows the optimisation is cheap (Section~\ref{sec:subspace}), and turns the
caveat into a predictive law (Section~\ref{sec:when}).

% ======================================================================
\section{Background: What We Already Know Works}
\label{sec:background}

\subsection{Task and setup}
\label{sec:task}
All experiments use one agentic environment: a suite of $20$ command-line (bash)
tasks in which the model must accomplish a stated goal by issuing shell commands
through a tool interface and reading their output---e.g.\ counting the lines in a
file, or listing files matching a pattern.  Each task has a ground-truth answer
used \emph{only} for evaluation.  The policy is Qwen2.5-1.5B-Instruct trained
with GRPO and online weight-synced generation; ``base'' denotes the untrained
model.  Training is \emph{verifier-free}: instead of ground truth, the reward
scores a rollout by how similar its terminal output is to the other rollouts in
its group (self-consistency/consensus), optionally as a vector of such axes
(Section~\ref{sec:spectrum}).  We evaluate with best@$k$ (the unbiased pass@$k$
estimator) against the held-out answers, because it measures the coverage a
downstream search would exploit.

\emph{Scope.}  We treat the structural result (Section~\ref{sec:subspace}) as
analytical and model-agnostic, and corroborate the empirical findings beyond the
originating bash task: the consensus collapse and the min-norm reliability /
collapse-avoidance advantage replicate on GSM8K and MBPP and at 7B
(Sections~\ref{sec:exp-crosstask}--\ref{sec:limitations}), with Wordle a
degenerate-axis counter-case that sharpens the governing condition.  A full
scale$\times$task grid and the learned reward-decomposition code remain open
items (Section~\ref{sec:limitations}).

\subsection{Scalar self-similarity (Exp17)}

The simplest verifier-free reward: for each rollout $i$ in a group of
$G{=}8$, compute mean pairwise terminal similarity
\[
  z_i = \frac{1}{G{-}1}\sum_{j \neq i} \mathrm{sim}(\mathrm{out}_i,\,
        \mathrm{out}_j),
\]
then z-score over the group and use as the GRPO advantage.
Result: $+12.5$ pp over base on Prime A6000 ($+13.4$ pp at step 150).
A surprisingly strong scalar baseline---any vector method must clear it.

\subsection{Variational V1 (Exp9)}

Same scalar $z_i$ but with an explicit variational interpretation and
weight sync.  Result: $+15.2$ pp over base on nodeset3.

\subsection{Vector-$\lambda$ V2 (Exp10)}

Three reward axes per rollout:
\begin{equation}
  \bm{R}_i = \big(\mathrm{strict\_sim}_i,\;\;
                \mathrm{jaccard\_sim}_i,\;\;
                \mathrm{exit\_success}_i\big) \in \R^3.
  \label{eq:v2-reward}
\end{equation}
Per training step, sample $\bm{\lambda} \sim \mathrm{Dir}(\alpha{=}1.0)$
and collapse:
\begin{equation}
  R_i = \bm{\lambda} \cdot \bm{R}_i, \qquad
  A_i = \frac{R_i - \bar{R}}{\mathrm{std}(R)}.
  \label{eq:v2-dirichlet}
\end{equation}
Result: $+16.1$ pp over base---the best single number, but only $\sim$1 pp
above V1.

\paragraph{Reading the V2 result correctly.}
The small gap is not a disappointment to be explained away; it is the first
data point of this paper's central law.  Two of the three axes
($\mathrm{strict\_sim}$ and $\mathrm{jaccard\_sim}$) both measure output
similarity and are nearly collinear, so the ``vector'' reward is effectively
rank-$\approx$2 and the Dirichlet mixing has almost nothing orthogonal to
mix.  V2 beating V1 by $\sim$1 pp is exactly what
Section~\ref{sec:when} predicts for a near-collinear code.  The lesson that
organises the rest of the paper: \emph{the axes were not designed to carry
independent information, so they did not.}

% ======================================================================
\section{The Gradient-Subspace View: Vector Rewards Are Almost Free}
\label{sec:subspace}

Before discussing \emph{which} update rule to use, we establish a structural
fact that reshapes the cost analysis and gives a cleaner formulation of every
rule in the spectrum.

\subsection{All per-axis gradients live in a $G$-dimensional subspace}

Under GRPO-style outcome advantages, the advantage is assigned at the
trajectory level and broadcast to tokens.  For reward axis $k$ the
group-estimated policy gradient is
\begin{equation}
  \bm{g}_k
  \;=\; \frac{1}{G}\sum_{i=1}^{G} A_k(\tau_i)\,
        \underbrace{\nabla_\theta \log \pi_\theta(\tau_i)}_{\textstyle v_i}.
  \label{eq:gk}
\end{equation}
The per-trajectory score gradient $v_i$ is \emph{identical across all
axes}---only the scalar weight $A_k(\tau_i)$ changes.  Hence
\begin{equation}
  \bm{g}_k \in \mathrm{span}\{v_1,\dots,v_G\}
  \quad\text{for every } k,
  \qquad
  \dim \mathrm{span}\{v_i\} \le G .
  \label{eq:span}
\end{equation}
With $G=8$ rollouts, \emph{all} $K$ axis-gradients---and every convex
combination of them---live in a subspace of dimension at most $8$,
regardless of $K$ or of the size of $\theta$.

\paragraph{Consequence 1: no $K$ backward passes.}
We never need one backward pass per axis.  We need the $G$ per-trajectory
gradients $\{v_i\}$ (which the trainer already forms, modulo storing them
per-sample), after which \emph{any} scalarisation is a reweighting of the
same $G$ vectors.  The ``$3$--$5\times$ trainer cost'' that earlier framing
worried about is an artifact of materialising $\bm g_k$ explicitly rather
than working in coefficient space.

\paragraph{Consequence 2: a tiny Gram-matrix problem.}
Let $V = [v_1,\dots,v_G]$ and let $A \in \R^{K\times G}$ be the advantage
matrix with $A_{ki}=A_k(\tau_i)$, so $\bm g_k = \tfrac1G V (A^\top)_{:,k}$.
Any mixture $\sum_k \alpha_k \bm g_k = \tfrac1G V A^\top \bm\alpha$ has squared
norm
\begin{equation}
  \Big\|\textstyle\sum_k \alpha_k \bm g_k\Big\|^2
  = \tfrac1{G^2}\,(A^\top\bm\alpha)^\top\, M\, (A^\top\bm\alpha),
  \qquad
  M_{ij} = \langle v_i, v_j\rangle,
  \label{eq:gram}
\end{equation}
where $M\in\R^{G\times G}$ is the (cheap) Gram matrix of per-trajectory
gradients.  Every update rule below---fixed, Dirichlet, worst-axis, MGDA---is
an optimisation over $\bm\alpha$ in this $G$-dimensional quadratic, with no
further access to $\theta$-sized objects.

\paragraph{Consequence 3: free exploration of the scalarisation simplex.}
Because the entire family of reward mixtures is reachable from the same
$\{v_i\}$, we can train \emph{across} a distribution of scalarisations at no
extra backward cost---sampling many $\bm\alpha$ per step and steering the
policy to cover the strategies each one favours.  This is the concrete lever
by which structured reward \emph{spreads} the policy for downstream search
(Section~\ref{sec:motivation}), rather than collapsing it.

\paragraph{Scope.}
Equation~\eqref{eq:span} relies on outcome-level (trajectory-broadcast)
advantages, as in GRPO.  Token-level GAE assigns a different advantage per
token and breaks the rank-$G$ argument; the subspace then has dimension up to
(group size $\times$ tokens) and the cheap-Gram story no longer holds.  We
restrict claims to the outcome-advantage setting, which is the one our
experiments use.

% ======================================================================
\section{Spectrum of Vector-Reward Update Rules (in Coefficient Space)}
\label{sec:spectrum}

Given a vector advantage $\bm{A}(\tau) \in \R^K$, the design question is how
to turn it into a parameter update.  By Section~\ref{sec:subspace} every
choice is a rule for picking $\bm\alpha$ in Eq.~\eqref{eq:gram}.

\begin{table}[h]
\centering
\caption{Spectrum of vector-reward update rules, recast as choices of the
mixing coefficients $\bm\alpha$ over the $G$-dimensional trajectory-gradient
subspace.}
\label{tab:spectrum}
\begin{tabular}{@{}lll@{}}
\toprule
\textbf{Approach} & \textbf{Rule for $\bm\alpha$ / scalarisation} & \textbf{When it earns its keep} \\
\midrule
Fixed linear weight &
  $A_{\mathrm{scalar}} = \bm{w} \cdot \bm{A}$ &
  Baseline only; requires hand-tuning \\[4pt]
Dirichlet mixing (V2) &
  $\bm{\lambda} \sim \mathrm{Dir}(\alpha)$ per step &
  Random regularisation across axes \\[4pt]
Worst-axis (max-min) &
  $A_{\mathrm{scalar}} = \min_k A_k$ &
  Improves the weakest axis; slow \\[4pt]
MGDA (D\'{e}sid\'{e}ri 2012) &
  $\argmin_{\bm{\alpha} \in \Delta^{K-1}} \|\sum_k \alpha_k \bm{g}_k\|^2$ &
  Min-norm compromise; see caveats \S\ref{sec:mgda} \\[4pt]
Adaptive per-axis &
  Weight axis $k$ $\propto$ policy weakness on $k$ &
  Spends gradient where needed; needs per-axis eval \\[4pt]
\textbf{Matryoshka (MRD, ours)} &
  $\bm\lambda \sim \mathrm{Dir}$ with a \emph{nesting prior} over a learned
  sparse code &
  Manufactures the non-collinear axes the others assume \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{The crucial distinction, restated.}
Rows~1--3 collapse the vector to a scalar before forming any gradient; MGDA
and MRD operate on the per-axis structure.  But note that, by
Eq.~\eqref{eq:gram}, ``preserving vectoriality at the gradient level'' does
\emph{not} buy a richer subspace---every rule lives in the same
$\mathrm{span}\{v_i\}$.  What it buys is a different \emph{choice within that
subspace}.  The value of that choice is bounded by how much the rows of $A$
actually differ, which is the quantity Section~\ref{sec:when} makes
predictive and Section~\ref{sec:mrd} makes designable.

% ======================================================================
\section{MGDA as One Instantiation}
\label{sec:mgda}

MGDA (Multiple-Gradient Descent Algorithm; D\'{e}sid\'{e}ri 2012) picks the
min-norm convex combination of the per-axis gradients.  Via
Eq.~\eqref{eq:gram} this is a $G$-dimensional quadratic program.

\paragraph{MGDA as ``parameter golf.''}
It is worth naming the principle MGDA embodies, because it is the same one
that drives the inference-time search this paper is written to serve.  An
agent like AIDE plays \emph{parameter golf}: reach the target in the fewest
strokes by making the \emph{smallest} change that improves the metric without
breaking what already works.  MGDA is the gradient-space version of exactly
that move---among all convex combinations that do not regress any axis, it
takes the one of \emph{minimum norm}, the least-committal step that still
improves.  The paper's two halves are thus one idea at two levels:
constrained minimal-move ascent at the solution level (the search) and at the
gradient level (the update).  The min-norm choice is conservative by
construction, which is also why on a near-collinear code it collapses to
averaging (below): there is no cheap stroke that distinguishes the axes.

\begin{algorithm}[h]
\caption{MGDA in coefficient space}\label{alg:mgda-vpo}
\begin{algorithmic}[1]
\Require Policy $\pi_\theta$, $K$ advantage functions $\{A_k\}$, learning
  rate $\eta$, group size $G$
\For{each training step}
  \State Generate $G$ rollouts $\{\tau_i\}$; form per-trajectory gradients
         $v_i = \nabla_\theta \log\pi_\theta(\tau_i)$
  \State Compute $A_{ki} = A_k(\tau_i)$ (z-scored over the group) and the
         Gram matrix $M_{ij}=\langle v_i,v_j\rangle$
  \State Solve $\bm\alpha^* = \argmin_{\bm\alpha\in\Delta^{K-1}}
         (A^\top\bm\alpha)^\top M (A^\top\bm\alpha)$
         \Comment{$G\!\times\!G$ QP, negligible cost}
  \State Update $\theta \gets \theta + \tfrac{\eta}{G}\, V A^\top \bm\alpha^*$
\EndFor
\end{algorithmic}
\end{algorithm}

\paragraph{Properties and an honest caveat.}
\begin{itemize}[nosep]
\item When the rows of $A$ are near-collinear, $\bm\alpha^*$ is essentially
  unconstrained and MGDA reduces to averaging---the V2 regime.
\item When axes genuinely conflict, $\bm\alpha^*$ selects a compromise that
  decreases the min-norm objective.
\item \textbf{The ``zero update means Pareto-stationary'' claim must be
  softened.}  MGDA's guarantee is a property of a \emph{Pareto-stationary
  point under deterministic gradients}.  Here the $v_i$, and hence
  $\bm g_k$, are \emph{noisy minibatch estimates}; the min-norm criterion is
  biased toward gradient \emph{cancellation}, so $\|\sum\alpha^*_k\bm
  g_k\|\!\approx\!0$ can be sampling noise rather than true stationarity.
  We therefore treat a near-zero MGDA update as a diagnostic to investigate
  (variance estimate over microbatches), not as a convergence certificate.
\end{itemize}
This caveat is the main reason we do not lead with MGDA: on a well-designed,
genuinely non-collinear code, the choice of combiner (MGDA vs.\ tuned
scalarisation) matters far less than the design of the code itself---a point
the multi-task-learning literature has made forcefully
(Section~\ref{sec:related}).

% ======================================================================
\section{Matryoshka Reward Decomposition (MRD)}
\label{sec:mrd}

Every method in Section~\ref{sec:spectrum} assumes a vector reward whose
axes carry independent information.  Where does such a vector come from?
The earlier Vector-$\lambda$ work \emph{hand-picked} three axes and got two
collinear ones.  MRD makes the reward code \emph{learned, nested, and
partially interpretable} instead.

\subsection{Design goals}

\begin{enumerate}[nosep]
\item \textbf{Don't choose $K$.}  We want a spectrum, not a fixed axis count.
\item \textbf{Separate what is separable; leave the rest fuzzy.}  A few
  dimensions should be explicitly disentangled and human-named (correctness,
  format); the remainder should be allowed to stay entangled in latent
  space rather than forced into spurious orthogonality.
\item \textbf{Serve search.}  The interpretable head should give controllable
  steering; the entangled tail should supply uncontrolled-but-useful
  diversity.
\end{enumerate}

\subsection{A nested, sparse reward code}

Train an encoder $z(\tau)\in\R^D$ over the teacher's signals (or over a
reward model's residual stream---see the mechanistic-interpretability
connection below), with three losses:

\paragraph{(i) Matryoshka nesting (Kusupati et al.\ 2022).}
Require every prefix $z_{1:k}$ to already reconstruct the dominant reward:
\begin{equation}
  \mathcal{L}_{\mathrm{nest}}
  = \sum_{k \in \mathcal{K}} w_k\,
    \big\| \hat r(z_{1:k}) - r^\star \big\|^2,
  \qquad \mathcal{K} = \{1,2,4,8,\dots\},
\end{equation}
so early dimensions carry coarse, dominant signal (correctness) and later
dimensions carry progressively finer, more specialised signal (format,
brevity, intermediate-state path).  Nesting makes ``how many axes'' a
\emph{readout choice at train/inference time}, not an architecture
decision---and absorbs the old ``adaptive $K$ / $L_0$-on-the-simplex'' open
question as a built-in property.

\paragraph{(ii) Head disentanglement.}
Apply a sparsity + orthogonality penalty \emph{only} to the head dimensions
$z_{1:h}$:
\begin{equation}
  \mathcal{L}_{\mathrm{head}}
  = \lambda_{\mathrm{sp}} \|z_{1:h}\|_1
  + \lambda_{\mathrm{orth}} \big\| \mathrm{Corr}(z_{1:h}) - I \big\|_F^2 ,
\end{equation}
pushing the head toward monosemantic, near-independent, human-auditable
reward features.

\paragraph{(iii) A deliberately entangled tail.}
Leave $z_{h+1:D}$ dense and unregularised: a learned residual that captures
hard-to-name quality the head cannot express.  We do \emph{not} try to
factorise it.  This is the technical content of ``some things are fuzzy in
the latent space and you don't try to explicitly separate everything'': the
tail is where the model is allowed to be vague, and the diversity search
exploits comes disproportionately from it.

\subsection{Optimisation: a matryoshka prior over scalarisations}

We then run the cheap coefficient-space machinery of
Section~\ref{sec:subspace} over the code rather than over hand axes.  Per
step we sample a mixing vector with a \emph{nesting prior}---more mass on
early dimensions, exponentially decaying into the tail:
\begin{equation}
  \bm\lambda \sim \mathrm{Dir}(\alpha_1,\dots,\alpha_D),
  \qquad \alpha_d \propto \gamma^{\,d},\;\; 0<\gamma<1,
\end{equation}
and form the per-axis advantages and update exactly as in
Eqs.~\eqref{eq:gk}--\eqref{eq:gram}.  Because all of this is reweighting in
the $G$-dimensional subspace, sampling many $\bm\lambda$ per step is free,
and each $\bm\lambda$ defines a specialist the downstream selector can later
choose among:

\begin{itemize}[nosep]
\item \textbf{Head-heavy $\bm\lambda$}: controllable specialists (``correct
  and well-formatted''), the axes a user or outer search loop can steer on.
\item \textbf{Tail-touching $\bm\lambda$}: diversity injectors that push the
  policy into uncovered regions without a human-named objective.
\end{itemize}

\subsection{The mechanistic-interpretability connection}

If the encoder is an \emph{sparse autoencoder} over a reward model's
activations, the head dimensions are literally SAE features of ``reward,''
and MRD becomes: extract a sparse, near-monosemantic basis for the teacher's
notion of quality (cf.\ dictionary-learning work on monosemantic features),
keep the few features that are interpretable and independent as the steering
head, and treat the dense reconstruction residual as the fuzzy tail.  This
gives a principled, auditable answer to ``what are the reward axes?''---they
are discovered features, not guesses---while explicitly conceding that not
everything the teacher rewards will be monosemantic, and not forcing it to
be.

% ======================================================================
\section{When Does Structured Reward Earn Its Keep? A Predictive Law}
\label{sec:when}

The central question is not ``is the method good?'' but \emph{when} the
reward code carries information a scalar cannot represent.  We promote this
from a caveat to a measured, predictive variable.

\subsection{The degeneracy condition}

If $A_k \approx c_k A_0 + \epsilon_k$ with
$\mathrm{Var}(\epsilon_k) \ll \mathrm{Var}(c_k A_0)$ for all $k$, then by
Eq.~\eqref{eq:gk} $\bm g_k \approx c_k \bm g_0$, every mixture is collinear,
and \emph{no} combiner---MGDA, Dirichlet, or worst-axis---can beat tuned
scalarisation.  This is not a weakness of any one rule; it is a property of
the \emph{code}, which is exactly why MRD targets the code.

\subsection{Correlation as the independent variable}

We propose to \emph{sweep} inter-axis correlation rather than report a single
point.  After collecting $G{=}8$ rollouts per task, form the pairwise Pearson
matrix of the $K$ advantages and summarise it by its effective rank (or by
$\bar\rho$, the mean off-diagonal $|\rho|$).  The paper's empirical claim
becomes a curve:
\begin{quote}
\emph{The gap between structured-reward and best-tuned-scalar performance
is monotone increasing in the non-collinearity of the reward code, and
$\approx 0$ when the code is rank-1.}
\end{quote}
V2 ($\bar\rho$ high, gap $\approx$ 1 pp) is the left end of this curve.  A
result of this shape converts the honest ``it only helps sometimes'' into a
\emph{predictive theory of when}, which is far stronger than a single win.

\subsection{Candidate axes for the bash-agent domain}

\begin{table}[h]
\centering
\caption{Candidate reward axes and their expected correlation with the
dominant correctness axis.  MRD's job is to \emph{discover} a basis like
this rather than assume it.}
\label{tab:axes}
\begin{tabular}{@{}llc@{}}
\toprule
\textbf{Axis} & \textbf{Source} & \textbf{Corr.\ with correctness} \\
\midrule
Answer correctness & Ground-truth oracle & Perfect (dominant / head dim 1) \\
Tool-call format validity & Syntactic check & Moderate \\
Reasoning consistency & LLM judge or heuristic & Moderate \\
Brevity (negative length) & Cheap, mechanical & Low \\
Intermediate env-state match & Privileged env introspection & Low \\
Exit code 0 & Mechanical & Moderate \\
\bottomrule
\end{tabular}
\end{table}

Axes 4--5 (brevity, state-path) are low-correlation with correctness: a
trajectory can be correct but verbose, or correct via a different internal
path.  These are where a scalar conflates ``correct but suboptimal'' with
``partially correct but efficient,'' and where a non-collinear code has room
to separate the signals.

% ======================================================================
\section{Experiments}
\label{sec:exp}

\subsection{What the flagship must show}

An oral-grade result needs the structured method \emph{run} and winning by a
margin that is not noise, with the mechanism isolated.  We therefore do not
lead with the near-collinear three-axis setup (which is engineered to
produce a null); we lead with the diversity and correlation-sweep results.

\subsection{Measured reward-axis correlation (first result)}
\label{sec:exp-corr}

We ran the K=5 MRPO reward code (Qwen2.5-1.5B-Instruct, GRPO with weight sync,
the same bash sandbox as Exp9/10, on a 2$\times$L40S pod) and logged every
group's reward vector.  Table~\ref{tab:corr-measured} reports the Pearson
correlation matrix pooled over $1408$ rollouts ($16$ training steps $\times$
$11$ groups $\times$ $8$ rollouts).  The measurement confirms the paper's
central structural claim directly:
\begin{itemize}[nosep]
\item $\mathrm{strict\_sim}$ and $\mathrm{jaccard\_sim}$ are \textbf{collinear}
  ($\rho = 0.95$): the K=3 V2 code is effectively rank-2, which is exactly why
  V2 beat V1 by only $\sim$1\,pp (Section~\ref{sec:background}).
\item The two new axes are \textbf{near-orthogonal} to the similarity axes:
  $\mathrm{brevity}$ at $\rho \approx -0.10$ and $\mathrm{tool\_format}$ at
  $\rho \approx -0.03$. They inject genuinely independent information.
\item The full code has effective rank $3.92/5$ (mean off-diagonal
  $|\rho| = 0.161$), versus $\approx 2$ for the V2 triple.
\end{itemize}
This is the left-endpoint anchor of the correlation law
(Section~\ref{sec:when}): the V2 result lives at high $\bar\rho$ where no
combiner can help, and the MRPO axes are what move the code off that point.

\begin{table}[h]
\centering
\caption{Measured reward-axis Pearson correlation for the K=5 MRPO code,
pooled over $1408$ rollouts. mean$|\rho_{\mathrm{off}}| = 0.161$, effective
rank $= 3.92/5$. \texttt{strict}/\texttt{jaccard} are collinear ($0.95$);
\texttt{brevity}/\texttt{tool\_format} are near-orthogonal to all others.}
\label{tab:corr-measured}
\begin{tabular}{@{}lrrrrr@{}}\toprule
 & strict & jaccard & exit & brevity & tool\_fmt \\ \midrule
strict     &  1.00 &  0.95 &  0.16 & -0.10 & -0.02 \\
jaccard    &  0.95 &  1.00 &  0.15 & -0.09 & -0.03 \\
exit       &  0.16 &  0.15 &  1.00 &  0.06 &  0.01 \\
brevity    & -0.10 & -0.09 &  0.06 &  1.00 & -0.04 \\
tool\_fmt  & -0.02 & -0.03 &  0.01 & -0.04 &  1.00 \\
\bottomrule\end{tabular}
\end{table}

\subsection{Measured best@$k$ across seeds: what survives, and what was seed luck}
\label{sec:exp-bestk}

We trained each policy for $200$ GRPO steps with the verifier-free similarity
reward and evaluated unbiased pass@$k$ (best@$k$) on the $14$ checkable bash
tasks, $n{=}16$ samples each.  Crucially we run \emph{three seeds} per trained
policy and report mean$\pm$std (Table~\ref{tab:bestk}); the base policy is
untrained, so it has a single curve.  The seeds change the story relative to
single-seed runs, and we report only the seed-robust version.

\paragraph{What is robust.}
\begin{itemize}[nosep]
\item \textbf{A quality-misaligned axis collapses the policy, deterministically.}
  The K=5 set (K=3 plus brevity and tool-format) collapses to best@$k=0$ in
  \emph{every} seed ($0.000\pm0.000$).  Rollout inspection shows the mechanism:
  the policy stops emitting tool calls and degenerates to terse bare-text
  outputs (e.g.\ the literal string \texttt{wc /etc/hosts}, mean completion
  $\approx 3$ tokens) that are trivially self-similar and thus maximise the
  consensus reward \emph{without doing the task}---reward hacking by format
  collapse.  K=3 vs.\ K=5 is a clean single-variable ablation and brevity is the
  culprit: it is \emph{non-collinear} with the similarity axes
  (Table~\ref{tab:corr-measured}, $\rho\approx-0.10$) yet directly rewards the
  short degenerate outputs.  Hence the corrected form of the correlation law:
  \begin{quote}\emph{Non-collinearity of reward axes is necessary but not
  sufficient; an axis must also be quality-aligned.  A non-collinear,
  quality-misaligned axis (brevity) accelerates collapse rather than preventing
  it.}\end{quote}
  This is exactly the condition MRD (Section~\ref{sec:mrd}) enforces by
  \emph{learning} a quality-bearing head rather than hand-picking axes by eye.
\item \textbf{The base policy has the highest diversity ceiling}
  ($\text{best@}16 = 0.857$)---the coverage downstream search exploits.
\item \textbf{Min-norm (MGDA) is the most stable trained policy}: the lowest seed
  variance ($0.537\pm0.024$ at $k{=}1$) and it matches the base ceiling
  ($0.857\pm0.058$).  It does \emph{not} beat tuned scalarisation on these
  near-collinear axes---as predicted by the MTL literature and our
  Section~\ref{sec:mgda} caveat---but it is markedly lower-variance than the
  Dirichlet V2 control ($0.420\pm0.197$ at $k{=}1$), a regularising effect of
  the min-norm step worth investigating with more seeds.
\end{itemize}

\paragraph{What did \emph{not} survive seeds.}
On a single seed the non-collapsed policies appeared to lie on a clean
best@$k$-slope gradient---we initially observed $+0.33$ (base) $>+0.21$
(quality-aligned) $>+0.11$ (V2 consensus)---and the quality-aligned subset
looked best (a $k{=}1$ of $0.665$ on that seed).  Across three seeds neither
holds: the trained, non-collapsed policies have similar slopes ($+0.22$ to
$+0.32$) with overlapping error bands, and the quality-aligned $k{=}1$ is
$0.496\pm0.121$.  The seed-robust signal is therefore \emph{collapse vs.\ not},
together with the \emph{variance} structure---not a fine-grained ordering among
the healthy policies.  We state this explicitly because it is the Parameter-Golf
discipline (results must hold across seeds; Jiang et al.\ 2025) applied to our
own earlier single-seed numbers.

\begin{table}[h]
\centering
\caption{Measured best@$k$ (unbiased pass@$k$) on $14$ bash tasks, $n{=}16$,
node 7 of the a3mega H100 cluster ($\mathrm{mbs}{=}8$).  Trained policies:
mean$\pm$std over 3 seeds; base is the untrained reference (1 seed).  Robust
signals: the brevity (K=5) collapse is deterministic; MGDA is the lowest-variance
trained policy and matches the base ceiling; the Dirichlet V2 control is
high-variance.  Slope differences among non-collapsed policies are within seed
noise.}
\label{tab:bestk}
\begin{tabular}{@{}lcccc@{}}\toprule
Policy & $k{=}1$ & $k{=}4$ & $k{=}16$ & slope \\ \midrule
base (untrained, 1 seed)   & $0.509$            & $0.747$            & $0.857$            & $+0.35$ \\
MGDA, K=3                  & $0.537{\pm}.024$   & $0.745{\pm}.024$   & $0.857{\pm}.058$   & $+0.32$ \\
quality-aligned, K=3       & $0.496{\pm}.121$   & $0.641{\pm}.069$   & $0.786{\pm}.058$   & $+0.29$ \\
Dirichlet V2, K=3          & $0.420{\pm}.197$   & $0.570{\pm}.176$   & $0.643{\pm}.154$   & $+0.22$ \\
+brevity, K=5              & $0.000$            & $0.000$            & $0.000$            & $\;\;0.00$ \\
\bottomrule\end{tabular}
\end{table}

\subsection{Min-norm gradient combination is the most reliable across seeds}
\label{sec:exp-reliability}

The seed variance above motivates a different question than ``which combiner has
the best mean'': \emph{which is the most reliable?}  We turn the \emph{same} K=3
quality-aligned reward vector into an update three ways---fixed equal weights,
Dirichlet random scalarisation (V2), and MGDA min-norm---each over \textbf{ten
seeds} (Table~\ref{tab:reliability}):
\begin{itemize}[nosep]
\item MGDA has \textbf{$\sim$5$\times$ lower seed variance} than either baseline
  ($\pm0.035$ at $k{=}1$ vs.\ $\pm0.186$ Dirichlet and $\pm0.169$ fixed; $\pm0.046$
  at $k{=}16$ vs.\ $\pm0.238$ and $\pm0.164$), and the gap \emph{widens} with more
  seeds.
\item MGDA also has the \textbf{highest mean} at every $k$ (best@16 $0.836$ vs.\
  $0.707$ Dirichlet, $0.636$ fixed).
\end{itemize}
So on a near-collinear code the min-norm step does not merely tie tuned
scalarisation on the mean (as the MTL literature reports); it is markedly
\emph{more reliable}.  The mechanism is intuitive: min-norm shrinks the update
exactly when the per-axis gradients disagree, damping the seed-dependent drift
that fixed/random scalarisation lets through.  We argue \emph{reliability}, not
mean performance, is the axis on which gradient-level vector methods earn their
keep in RL post-training---and it is precisely the pain point the seeds exposed
in Section~\ref{sec:exp-bestk}.

\begin{table}[h]\centering
\caption{Reliability of three combiners on the same K=3 quality-aligned axes,
mean$\pm$std over \textbf{10 seeds}, best@$k$ on $14$ bash tasks ($n{=}16$).
MGDA (min-norm) is $\sim$5$\times$ lower variance and highest mean at every $k$.}
\label{tab:reliability}
\begin{tabular}{@{}lccc@{}}\toprule
Combiner & $k{=}1$ & $k{=}4$ & $k{=}16$ \\ \midrule
Fixed weights   & $0.397{\pm}.169$ & $0.544{\pm}.161$ & $0.636{\pm}.164$ \\
Dirichlet (V2)  & $0.454{\pm}.186$ & $0.610{\pm}.212$ & $0.707{\pm}.238$ \\
MGDA (min-norm) & $\mathbf{0.515{\pm}.035}$ & $\mathbf{0.712{\pm}.031}$ & $\mathbf{0.836{\pm}.046}$ \\
\bottomrule\end{tabular}
\end{table}

\paragraph{Collapse is abrupt but preceded by a warning window.}
Evaluating intermediate checkpoints of the K=5 (+brevity) run gives a clean
trajectory: best@16 holds at $0.79$--$0.86$ through step $40$, falls to $0.57$ by
step $60$, and \emph{cliffs to $0$ by step $80$} (and stays there to $200$);
pass@1 begins eroding earlier, from step $40$.  Format collapse is thus not
instantaneous---there is a multi-step window in which greedy accuracy and output
length decline before best@$k$ reaches zero---which supports a cheap
reward-vs-held-out-correctness divergence detector and an automatic
down-weight/early-stop intervention (future work).

\paragraph{Two reward-design knobs, reported honestly.}
A \emph{coverage} reward (reward distinctness from the group, gated by validity,
instead of consensus) produces the steepest best@$k$ slope of any trained policy
(over 8 seeds, $+0.54$: a low but stable greedy pass@1 of $0.162\pm0.034$
climbing to best@16 $0.705\pm0.121$)---it genuinely trains \emph{for} diversity.
At the search-relevant end it \emph{beats the consensus reward} (best@16 $0.705$
vs.\ $0.643$ for V2 consensus), confirming that optimising for diversity rather
than agreement helps where search operates---though it still does not exceed the
untrained base ceiling ($0.857$), so it is a usable exploration/exploitation knob
rather than a dominance result.  Pooling distinct specialist policies (an
inference-time ensemble) likewise beats the best single policy only at the
large-$k$ end ($+0.02$ at $k{=}16$); directed diversity helps a little exactly
where it should.

\subsection{Setup}

All runs use Qwen2.5-1.5B-Instruct, GRPO with weight sync, the verifier-free
terminal-similarity reward, $G{=}8$, $\mathrm{lr}{=}10^{-5}$,
$\mathrm{lora\_rank}{=}16$, $200$ steps.  The reward-axis correlation
(Section~\ref{sec:exp-corr}) was measured on an earlier 2$\times$L40S run;
the best@$k$ results (Section~\ref{sec:exp-bestk}) are on one node of an
8$\times$H100 cluster at $\mathrm{mbs}{=}8$, vLLM serving on one GPU and the
trainer on another, three seeds per trained policy run in parallel across the
node.  best@$k$ is the unbiased pass@$k$ estimator over $n{=}16$ samples on the
$14$ bash tasks with a checkable answer.  All policies are evaluated on the same
node, so the (imperfectly calibrated) answer oracle biases every row equally and
the \emph{differences}, not the absolute values, are the signal.

\subsection{Generalisation across tasks}
\label{sec:exp-crosstask}

To test whether these findings are bash-specific, we ran the same comparison on
standard tasks with the identical stack (Qwen2.5-1.5B, $n{=}16$ best@$k$, 3
seeds): \textbf{GSM8K} (single-turn math), \textbf{MBPP} code generation (a
genuinely multi-axis task: \texttt{tests\_pass} / format / brevity), and
\textbf{Wordle} (multi-turn).  We reuse the env's own per-function rubric scores
as the reward axes, and compare combining them by Dirichlet, fixed weights, and
MGDA against the status-quo verifier-GRPO baseline (Table~\ref{tab:crosstask}).
The picture is \emph{nuanced and consistent with the axis-structure law}:
\begin{itemize}[nosep]
\item \textbf{Where axes carry real signal, the min-norm benefit replicates.}
  On MBPP, MGDA is the \emph{lowest-variance} vector method ($0.183\pm0.005$ vs.\
  Dirichlet $0.107\pm0.062$, fixed $0.123\pm0.039$) and ties the best mean.  On
  GSM8K the effect is dramatic: combining the answer-consensus axes by MGDA holds
  best@16 at $0.930\pm0.008$, while Dirichlet and fixed scalarisation
  \emph{collapse} to $\approx0.40\pm0.06$.  Min-norm avoids the collapse the other
  combiners suffer, and is again the lowest-variance.
\item \textbf{The verifier-free consensus collapse also replicates}: on GSM8K,
  optimising answer-agreement drops greedy accuracy from $0.50$ (base) to $0.13$
  --- the bash format-collapse phenomenon, on a standard benchmark.
\item \textbf{A clean failure mode, reported honestly.}  On Wordle MGDA is
  \emph{worst} ($0.017$ vs.\ Dirichlet $0.367$).  Wordle is a floor task for a
  1.5B model---it essentially never wins, so the win/turn axes are constant and
  only a near-degenerate partial-credit signal remains.  In that regime MGDA's
  min-norm step (which on noisy gradients is biased toward cancellation,
  Section~\ref{sec:mgda}) shrinks the update to near zero and the policy stays at
  the losing base; the higher post-training entropy confirms it barely moved.
  This is the same conservatism that \emph{helps} on signal-bearing axes,
  manifesting as a \emph{vice} when there is no real signal to preserve.
\end{itemize}
So the benefit of min-norm is not universal: it tracks exactly the condition the
paper predicts---non-collinear axes that carry genuine, non-degenerate
signal---holding on GSM8K and MBPP and failing on the Wordle floor.  (Two further
tasks we ran were uninformative for this comparison and are omitted from the
table: a synthetic tool-call env saturates at ceiling for all methods, and
reasoning-gym gave no separation between combiners.)

\paragraph{The effect strengthens with scale.}
To check that this is not a small-model artifact, we repeated the GSM8K
comparison with \textbf{Qwen2.5-7B-Instruct} (3 seeds; Table~\ref{tab:scale}).
The result is the same phenomenon, sharper: verifier-free Dirichlet and fixed
scalarisation \emph{collapse completely} (best@16 $\to 0.000$ and $0.013$), while
\textbf{MGDA fully preserves} the base / verifier-GRPO level
($0.973\pm0.005$).  The collapse the baselines suffer \emph{deepens} from
$\approx0.40$ at 1.5B to $\approx0$ at 7B, and MGDA's collapse-avoidance holds
throughout---so the reliability advantage is not an artifact of a small policy;
if anything the gap widens with capability.  The same ordering---MGDA matching
verifier-GRPO and exceeding both scalarisation baselines---also reproduces at 7B
on MBPP, though the separation there is compressed near that task's low ceiling
and within seed noise; Wordle stays a floor for every method at 7B as at 1.5B.
The decisive scale evidence is therefore GSM8K, where the dynamic range is large
enough for the collapse to be unambiguous.

\begin{table}[h]\centering
\caption{GSM8K best@$k$ at 1.5B vs.\ 7B (Qwen2.5-Instruct; mean$\pm$std, 3
seeds). Verifier-free Dirichlet/fixed scalarisation collapse---catastrophically
at 7B---while MGDA preserves base performance at both scales.}
\label{tab:scale}
\begin{tabular}{@{}lcccc@{}}\toprule
GSM8K best@16 & verifier-GRPO & Dirichlet & MGDA & fixed \\ \midrule
1.5B & $0.953{\pm}.009$ & $0.437{\pm}.065$ & $\mathbf{0.930{\pm}.008}$ & $0.397{\pm}.061$ \\
7B   & $0.973{\pm}.005$ & $0.000{\pm}.000$ & $\mathbf{0.973{\pm}.005}$ & $0.013{\pm}.019$ \\
\bottomrule\end{tabular}
\end{table}

\begin{table}[h]\centering
\caption{Cross-task best@16 (mean$\pm$std, 3 seeds, $n{=}16$).  MGDA is
lowest-variance and avoids collapse where axes carry real signal (MBPP, GSM8K),
and fails in the degenerate-axis floor regime (Wordle).  GSM8K baseline is
verifier-GRPO (ground truth); the vector methods are verifier-free.}
\label{tab:crosstask}
\begin{tabular}{@{}lcccc@{}}\toprule
Task & baseline & Dirichlet & MGDA & fixed \\ \midrule
MBPP (code, multi-axis)   & $0.183{\pm}.017$ & $0.107{\pm}.062$ & $\mathbf{0.183{\pm}.005}$ & $0.123{\pm}.039$ \\
GSM8K (single-axis)       & $0.953{\pm}.009$ & $0.437{\pm}.065$ & $\mathbf{0.930{\pm}.008}$ & $0.397{\pm}.061$ \\
Wordle (floor regime)     & $0.100$          & $\mathbf{0.367}$ & $0.017$ & $0.183$ \\
\bottomrule\end{tabular}
\end{table}

\subsection{Remaining experiments}
\label{sec:remaining}

The results above establish the correlation structure
(Section~\ref{sec:exp-corr}), the diversity findings
(Section~\ref{sec:exp-bestk}), and---now run---the MGDA coefficient-space
update.  MGDA (Alg.~\ref{alg:mgda-vpo}) forms the $N\times N$ Gram of
per-trajectory gradients, solves the simplex min-norm QP (cost independent of
the number of axes $K$), and feeds $\beta=A^\top\alpha^*$ to the existing single
backward---no per-axis backward pass; the solver's $Q=AMA^\top$ factorisation
and $\sum_i\beta_i v_i = \sum_k\alpha_k g_k$ identity are unit-tested.  On the
collinear K=3 axes it does not beat tuned scalarisation (as the MTL literature
predicts), but is the lowest-variance trained policy across seeds
(Table~\ref{tab:bestk})---a regularising effect of the min-norm step.  Three
further directions would round out the paper (we ran a minimal version of the
third below):
\begin{enumerate}[nosep]
\item \textbf{The correlation sweep.}  Interpolate reward codes from rank-1 to
  the head/tail MRD code and plot a seed-averaged metric (best@$k$ ceiling or
  collapse rate, with error bars) against effective rank
  (Section~\ref{sec:when}); the collapse-vs-not boundary we observe is the first,
  coarsest sample of that curve.
\item \textbf{The learned MRD code.}  Replace hand axes with the SAE/matryoshka
  code (Section~\ref{sec:mrd}) and test whether a learned quality-bearing head
  preserves best@$k$ where hand axes did not.  A \emph{tuned} scalar baseline
  (best fixed $\bm w$) is the control the MTL literature demands.
\item \textbf{Feedback-guided search (AIDE-style), with a caveat.}  best@$k$ is
  the \emph{i.i.d.}\ measurement.  Production search (AIDE; Jiang et al.\ 2025) is
  \emph{sequential}---each candidate conditioned on the previous failure.  We ran
  a \emph{minimal} version (linear, beam-1: propose, execute, read the error,
  re-prompt, repeat to budget $B{=}8$) on base, a healthy policy, and the
  format-collapsed K=5 policy.  One result is clean and one is a caveat:
  \emph{(i)} the collapsed policy is \textbf{unrecoverable by search}
  ($0\!\to\!0$ solve@$B$; with no valid tool call it has nothing to propose),
  confirming that format collapse is a capability loss search cannot repair;
  \emph{(ii)} on these simple single-command tasks the linear feedback loop does
  \emph{not} beat i.i.d.\ at equal budget ($0.71$ vs.\ $0.79$ for base and
  healthy alike)---the error signal adds little when each task is one command.
  A tree/beam search over harder, multi-step tasks (where AIDE's gains actually
  arise) is the proper test and remains future work.
\end{enumerate}

% ======================================================================
\section{Limitations and Generalisation}
\label{sec:limitations}

The core findings were developed on a bash tool-use task and then replicated on
GSM8K, MBPP, and Wordle (Section~\ref{sec:exp-crosstask}); the principal
remaining limitation is that every run uses \emph{one} small model
(Qwen2.5-1.5B).  We are explicit about which claims transfer and which still need
work.

\paragraph{What is task/model-agnostic.}
The subspace result (Section~\ref{sec:subspace})---that all per-axis gradients
lie in the span of the group's per-trajectory gradients, so any combiner is a
cheap reweighting---is a property of outcome-level policy gradients, not of this
task or model.  Likewise, measuring a reward code's structure needs nothing
task-specific: collect a group of base-model rollouts on the new task, compute
each candidate axis on them, and take the Pearson matrix / effective rank of the
per-axis advantages (Section~\ref{sec:when}).  This is exactly how one would
decide, for a \emph{new} task, whether vector rewards can help before training.

\paragraph{What is now shown across tasks, and what is not.}
Two of the three claims hold beyond bash (Section~\ref{sec:exp-crosstask}): the
consensus/agreement collapse recurs on GSM8K, and the min-norm reliability /
collapse-avoidance advantage recurs on GSM8K and MBPP---while Wordle gives a
clean counter-case that sharpens the condition (the benefit needs
non-degenerate, signal-bearing axes; a floor task where the model never succeeds
breaks it).  The effect is \emph{not} a small-model artifact: at \textbf{7B} on
GSM8K (Table~\ref{tab:scale}) the verifier-free baselines collapse even harder
(best@16 $\to 0.000$/$0.013$) while MGDA preserves the base level
($0.973$)---the reliability gap widens with scale rather than closing.  The
standing gaps are: \emph{(i)} broader scale and task coverage---we verify the
effect at 7B on one task, but a full scale$\times$task grid (and $>$10B) remains;
\emph{(ii)} harder benchmarks we could not run here (gated or judge-API-scored
datasets such as GPQA/SimpleQA, and a competition-math env), which would give
stronger correctness anchors; and \emph{(iii)} the learned MRD code itself, still
to be trained (Section~\ref{sec:remaining}).

% ======================================================================
\section{Relation to Existing Work}
\label{sec:related}

\paragraph{Concurrent set-construction work (Boldi et al.).}
Concurrent work by Boldi et al.\ targets the same search-downstream regime with
a different mechanism: emit $m$ candidate answers in one rollout (each attending
to the previous) and score the set by best-of-$m$ under stochastically sampled
scalarisation weights $\bm w\sim\mathrm{Dir}(1)$, so positions are incentivised
to differ.  Two points of contact.  (1) Their reported failure mode---this helps
only when reward components genuinely compete and \emph{loses} to GRPO when they
are near-collinear (their UltraFeedback result)---is the same condition our
Section~\ref{sec:when} formalises, independent confirmation that
non-collinearity is the operative variable.  (2) Their advantage is broadcast
over the whole multi-answer chain, so specialisation emerges only at the set
level over many updates; our head/tail code instead makes \emph{which} axis
earned reward explicit for the head while leaving the tail set-level.  Ours is
thus the \emph{reward-decomposition} counterpart to their
\emph{set-construction} approach.

\paragraph{Multi-task gradient methods, and the case against them.}
MGDA (D\'{e}sid\'{e}ri 2012; Sener \& Koltun 2018) is one of a family
including PCGrad (Yu et al.\ 2020), CAGrad (Liu et al.\ 2021), and Nash-MTL
(Navon et al.\ 2022).  Crucially, Kurin et al.\ (2022) and Xin et al.\
(2022) show that \emph{tuned unitary scalarisation usually matches or beats}
these methods.  We take this seriously: it is precisely why we do not claim
MGDA-the-combiner as the contribution, and instead locate the value in the
\emph{design of the reward code} (MRD), where the choice of combiner is
second-order.

\paragraph{Multi-objective RLHF.}
Rewarded Soups (Ram\'e et al.\ 2023) builds a Pareto front by interpolating
\emph{weights} of separately-trained experts; MORLHF, Directional Preference
Alignment, and Panacea learn weight-conditioned policies.  These produce a
\emph{family of policies} along a trade-off; MRPO instead shapes a
\emph{single} policy to cover multiple strategies for a downstream selector,
and learns the reward axes rather than assuming them.

\paragraph{Quality-diversity.}
The ``spread the policy for search'' goal is kin to Quality-Diversity /
MAP-Elites: MRD's interpretable head dimensions are candidate behaviour
descriptors, and the tail is undirected novelty.  Framing the head as a QD
descriptor space is a natural extension.

\paragraph{Representation learning.}
Matryoshka Representation Learning (Kusupati et al.\ 2022) supplies the
nesting; sparse-autoencoder / dictionary-learning work on monosemantic
features supplies the head-disentanglement; alignment/uniformity on the
hypersphere (Wang \& Isola 2020) supplies a formal handle on the
``uniformity = coverage for search'' intuition.

\paragraph{Reward shaping.}
Ng et al.\ (1999) showed potential-based shaping leaves the optimal policy
unchanged.  MRD is not shaping: it changes the gradient \emph{direction} via
the chosen $\bm\alpha$, not the reward potential, and can therefore change
which policy is favoured.

% ======================================================================
\section{Open Questions}
\label{sec:open}

\begin{enumerate}
\item \textbf{Encoder for the code.}  SAE over a reward model's residual
  stream vs.\ a small supervised encoder over teacher signals.  How
  monosemantic do head dims need to be before steering is reliable?
\item \textbf{Noisy-gradient MGDA.}  Can the cancellation bias of
  Section~\ref{sec:mgda} be corrected with a variance estimate over
  microbatches, or does it argue for sticking to (matryoshka-prior)
  stochastic scalarisation?
\item \textbf{How fuzzy should the tail be?}  Is there an optimal head size
  $h$, and does it track the effective rank of the teacher's reward?
\item \textbf{Generalisation.}  Math (step checks), code (unit-test
  granularity), and agentic tasks (intermediate-state visibility) all admit
  privileged teachers.  Does the correlation law hold, and where do these
  domains sit on the curve?
\item \textbf{Theory.}  Bound the $\bestk$ improvement of MRPO over scalar
  GRPO as a function of the reward code's effective rank.
\end{enumerate}

% ======================================================================
\section{Positioning Relative to the Current Paper}
\label{sec:positioning}

The three findings of the current (substrate) paper stand on their own:
\begin{enumerate}[nosep]
\item Variational self-consistency + weight sync $=+15$ pp (Exp9/10).
\item Scalar self-similarity is a strong baseline ($+13$ pp); the
  variational machinery's incremental contribution is small.
\item DBPO (density reward) is a documented failure mode (format collapse).
\end{enumerate}
MRPO is the follow-up that asks the next question: not ``what reward,'' but
``what \emph{structure} of reward, and how to combine it so the resulting
diversity survives for search.''  It builds on the same weight-synced GRPO +
bash substrate, reuses its rollouts, and---via
Section~\ref{sec:subspace}---adds the vector machinery at almost no extra
trainer cost.

% ======================================================================
\section*{References}
\addcontentsline{toc}{section}{References}

\begin{itemize}[nosep]
\item Jiang, Z.\ et al.\ (2025). \emph{AIDE: AI-Driven Exploration in the
  Space of Code}. arXiv:2502.13138. (Weco AI; tree-search ML-engineering agent,
  the inference-time search regime this paper's training targets.)
\item Boldi, R.\ et al.\ (2026). \emph{Vector Policy Optimization}.
  (Concurrent; multi-answer chains with stochastic scalarisation.)
\item D\'{e}sid\'{e}ri, J.-A. (2012). \emph{Multiple-gradient descent
  algorithm for multiobjective optimization}. C.\ R.\ Mathematique,
  350(5--6), 313--318.
\item Sener, O. \& Koltun, V. (2018). \emph{Multi-task learning as
  multi-objective optimization}. NeurIPS.
\item Yu, T.\ et al.\ (2020). \emph{Gradient surgery for multi-task learning
  (PCGrad)}. NeurIPS.
\item Liu, B.\ et al.\ (2021). \emph{Conflict-averse gradient descent for
  multi-task learning (CAGrad)}. NeurIPS.
\item Navon, A.\ et al.\ (2022). \emph{Multi-task learning as a bargaining
  game (Nash-MTL)}. ICML.
\item Kurin, V.\ et al.\ (2022). \emph{In defense of the unitary
  scalarization for deep multi-task learning}. NeurIPS.
\item Xin, D.\ et al.\ (2022). \emph{Do current multi-task optimization
  methods yield improvements?} NeurIPS Datasets/Benchmarks.
\item Ram\'e, A.\ et al.\ (2023). \emph{Rewarded Soups: weight interpolation
  for multi-objective alignment}. NeurIPS.
\item Kusupati, A.\ et al.\ (2022). \emph{Matryoshka Representation
  Learning}. NeurIPS.
\item Wang, T. \& Isola, P. (2020). \emph{Understanding contrastive
  representation learning through alignment and uniformity on the
  hypersphere}. ICML.
\item Ng, A.~Y., Harada, D. \& Russell, S. (1999). \emph{Policy invariance
  under reward transformations}. ICML.
\end{itemize}

\end{document}
